Rusty Linux: Advances in Rust for Linux Kernel Development

### 2024-06-06 paper.tex
Update on Overleaf.

For our research methodology, we followed the Kitchenham and Charters~\cite{Stuart2007-cc} methodology to conduct SLR in software engineering. We divided our review into three discrete phases, planning the review, conducting the review, and reporting the review results. The following sections detail our review process.

@@

To begin our study, we must first confirm the need for a SLR. The recent report released by
ONCD~\cite{United_States_Gov2024-pp} has conveniently done this job for us by compiling a report
detailing the need for research in the domain of memory safety. While the ONCD report detailed a two-pronged approach, for this reflection paper, we will be focusing on a memory-safe programming language,
specifically Rust. While there are many modern memory-safe programming languages available for
software developers to use, Rust is one of the few languages that is feasible to use when developing
operating system kernels due to its lack of runtime and garbage collector~\cite{Ojeda2021-ul}. Even the Linux kernel is
in the early stages of adding real support for
Rust~\cite{The_kernel_development_community_undated-iw} thus, we will focus on Rust as a primary
candidate to replace the aging C programming language.

@@

\begin{itemize}
    \item \textbf{RQ1:} What are the existing approaches for implementing operating system kernels in Rust?
    \item \textbf{RQ2:} What are the performance implications of using Rust for operating system kernel development?
      %in terms of throughput, latency, and resource utilization?
    \item \textbf{RQ3:} What are the major challenges and limitations when
      developing operating system kernels in Rust?% particularly in comparison to other languages?
    \item \textbf{RQ4:} What are the lessons learned when developing operating systems kernels in Rust?
\end{itemize}

@@

Once we had our initial collection of papers from the databases, we merged all duplicate papers into
one record and then started our first pass, which consisted of reading the title and abstract and
either marking the paper as \textit{include} or \textit{exclude}. Once the initial screening was
completed we were left with 26 potential papers that we needed an in-depth reading with an
additional 2 papers were found during forward and backward snowballing.

@@

In this section, we present our findings. We address each research question from \textbf{RQ1} - \textbf{RQ3}.

@@

The integration of Rust into operating system kernel development has garnered significant attention
due to Rust's promise of memory safety and concurrency guarantees. Several approaches and
methodologies have emerged to leverage Rust for kernel development, each addressing different
aspects of kernel functionality and integration.  We found several approaches that researchers are
taking to integrate Rust into kernel development. The first approach is a greenfield approach
where the operating system architecture is redesigned and built from the ground up in order to fully
leverage the Rust programming language. The second approach is the incremental integration of Rust
into existing kernels, such as the Linux
kernel~\cite{The_kernel_development_community_undated-iw}. This methodology involves gradually
rewriting or augmenting specific components of the kernel with Rust while maintaining compatibility
with the existing C codebase. Table~\ref{tab:RQ1} summarizes our findings regarding what existing
approaches researchers are using to integrate rust into the kernel.

For the first approach, Boos et al.~\cite{Boos2020-zh} created an experimental operating system
named Theseus that operates in a single address space and single privilege level and uses properties
of the Rust programming language to realize isolation instead of relying on hardware. This novel
approach pushes some of the roles that a traditional operating system would take onto the
compiler. The design of Theseus uses a novel cell-based structure where ownership of memory and
resources is enforced by the compiler, thus avoiding these checks during runtime to get improved
performance. In addition to memory safety, Theseus has a goal of addressing the state spill
problem. State spill happens when a single service in an operating system can harbor a state change
induced by interacting with other services. That state change can eventually cause a system crash or
system instability at some point in the future making it very difficult to track down the problem
due to the separation between when the problem occurred and when it was observed. Theseus OS is an
operating system that was designed to match the Rust language instead of the more traditional route
of matching the hardware.

While single address space operating systems present a unique opportunity to push more
responsibility into the compiler a more traditional micro-kernel implementation was taken by Chen X
et al.~\cite{Chen2023-wb} and Liang et al.~\cite{Liang2021-bo}. By pushing most kernel functionality
to user space, the authors limited the surface area that needs to be proved. Chen X et al.~\cite{Chen2023-wb} took
their implementation one step further by formally verifying their new micro-kernel named Atmosphere
by using both the liner type system of Rust in combination with a SMT solver. They were
able to get a 7.5:1 proof-to-code ratio, which is higher than other formerly verified micro-kernels
SeL4 and CeriKOS both have proof code ratios of 19:1 and 20:1, respectively. While this
approach was novel drivers running in userspace are still not formally verified and don't have to be
written in Rust.

Unikernels present an ideal surface for prototyping Rust~\cite{Madhavapeddy2014-zw}. Unikernels are designed to do one thing
and one thing only, thus limiting the complexity that a general-purpose operating system has to
deal with. This reduced complexity was explored by Lankes et al.~\cite{Lankes2019-cm}. The authors
detailed their experiences with rewriting an existing Unikernel, HermitCore, in Rust. The new
Unikernel named RustyHermit consists of only 3.27\% unsafe Rust, with the rest
of the code base consisting of safe Rust, dramatically decreasing the surface area where memory errors
could originate.

Finally, Li and Sato~\cite{Li2024-yb} explored using Rust to write an exokernel named W-Kernel. The authors
proposed a novel architecture for an OS by embedding a WebAssembly(WASM) runtime into the kernel that can run programs written in any language that can compile down to WASM.

>>>>

For our research methodology, we followed the Kitchenham and Charters~\cite{Stuart2007-cc} methodology to conduct SLR in software engineering. We divided our review into three discrete phases, planning the review, conducting the review, and reporting the review results. The following sections detail our review process and are diagrammed in figure~\ref{fig:process}.

 \begin{figure*}
    \centering
     \includegraphics[width=7in]{process-diagram.pdf}
     \caption{Process Diagram for Paper selection}
     \label{fig:process}
 \end{figure*}

@@

To begin our study, we must first confirm the need for a SLR. The recent report released by
ONCD~\cite{United_States_Gov2024-pp} has conveniently done this job for us by compiling a report
detailing the need for research in the domain of memory safety. While the ONCD report detailed a two-pronged approach, for this reflection paper, we will be focusing on a memory-safe programming language,
specifically Rust. While there are many modern memory-safe programming languages available for
software developers to use, Rust is one of the few languages that is feasible to use when developing
operating system kernels due to its lack of runtime and garbage collector~\cite{Ojeda2021-ul}. The Linux kernel is currently
in the early stages of adding real support for
Rust~\cite{The_kernel_development_community_undated-iw} thus, we will focus on Rust as a primary
candidate to replace the aging C programming language.

@@

\begin{itemize}
    \item \textbf{RQ1:} What are the existing approaches for implementing operating system kernels in Rust?

    All though Rust in the Linux kernel is very recent we would like to see what approaches researches have explored in any operating system. Experiential and research kernels can provide new ideas that can be integrated into the Linux kernel.
    \item \textbf{RQ2:} What are the performance implications of using Rust for operating system kernel development?

    The C programming language has the advantage of 40 years of improvements and lessons learned regarding how to write fast code. Rust being a new language still has many unexplored areas with how to write fast code especially in kernel space.
      %in terms of throughput, latency, and resource utilization?
    \item \textbf{RQ3:} What are the major challenges and limitations when
      developing operating system kernels in Rust?% particularly in comparison to other languages?

      Kernel space presents unique challenges for the software engineer using Rust. Rust was not designed to implement kernels and is more restrictive than C. Thus, this research question is looking for any stumbling blocks that have already been found.
    \item \textbf{RQ4:} What are the lessons learned when developing operating systems kernels in Rust?

    After the work has been completed we want to find all the research that reflects on what went well and what didn't.
\end{itemize}

@@

Once we had our initial collection of papers from the databases, we merged all duplicate papers into
one record and then started our first pass, which consisted of reading the title and abstract and
either marking the paper as \textit{include} or \textit{exclude}. Once the initial screening was
completed we were left with 26 potential papers that we needed an in-depth reading with an
additional 6 papers were found during forward and backward snowballing.

@@

In this section, we present our findings. We address each research question from \textbf{RQ1} - \textbf{RQ4}.

@@

The integration of Rust into operating system kernel development has garnered significant attention
due to Rust's promise of memory safety and concurrency guarantees. Several approaches and
methodologies have emerged to leverage Rust for kernel development, each addressing different
aspects of kernel functionality and integration.  We found several approaches that researchers are
taking to integrate Rust into kernel development. The first approach is a greenfield approach
where the operating system architecture is redesigned and built from the ground up in order to fully
leverage the Rust programming language~\cite{Culic2022-bk, Boos2020-zh}. The second approach is the incremental integration of Rust
into existing kernels, such as the Linux
kernel~\cite{The_kernel_development_community_undated-iw, Miller2021-pg, Oikawa2023-ms}. This methodology involves gradually
rewriting or augmenting specific components of the kernel with Rust while maintaining compatibility
with the existing C codebase. Finally, researches are porting existing C kernels over to Rust and then evaluating the differences. Table~\ref{tab:RQ1} summarizes our findings regarding what existing
approaches researchers are using to integrate rust into the kernel.

For the first approach, Boos et al.~\cite{Boos2020-zh} created an experimental operating system
named Theseus that operates in a single address space and single privilege level and uses properties
of the Rust programming language to realize isolation instead of relying on hardware. This novel
approach pushes some of the roles that a traditional operating system would take onto the
compiler. The design of Theseus uses a novel cell-based structure where ownership of memory and
resources is enforced by the compiler, thus avoiding these checks during runtime to get improved
performance. In addition to memory safety, Theseus has a goal of addressing the state spill
problem. State spill happens when a single service in an operating system can harbor a state change
induced by interacting with other services. That state change can eventually cause a system crash or
system instability at some point in the future making it very difficult to track down the problem
due to the separation between when the problem occurred and when it was observed. Theseus OS is an
operating system that was designed to match the Rust language instead of the more traditional route
of matching the hardware. Li and Sato~\cite{Li2024-yb} explored using Rust to write an exokernel named W-Kernel. The authors proposed a novel architecture for an OS by embedding a WebAssembly(WASM) runtime into the kernel that can run programs written in any language that can compile down to WASM.

For the second approach researches take a more  traditional micro-kernel implementation~\cite{Chen2023-wb, Liang2021-bo}. By pushing most kernel functionality
to user space, the authors limited the surface area that needs to be proved. Chen X et al.~\cite{Chen2023-wb} took
their implementation one step further by formally verifying their new micro-kernel named Atmosphere
by using both the liner type system of Rust in combination with a SMT solver. They were
able to get a 7.5:1 proof-to-code ratio, which is higher than other formerly verified micro-kernels
SeL4 and CeriKOS both have proof code ratios of 19:1 and 20:1, respectively. While this
approach was novel drivers running in userspace are still not formally verified and don't have to be
written in Rust.

Finally researchers have taken existing Unikernels written in C and ported them to to Rust. Unikernels are designed to do one thing
and one thing only, thus limiting the complexity that a general-purpose operating system has to
deal with~\cite{Madhavapeddy2014-zw}. This reduced complexity allowed  Lankes et al.~\cite{Lankes2019-cm} to take HermitCore and rewrite it in Rust. The new
Unikernel named RustyHermit consists of only 3.27\% unsafe Rust, with the rest of the code base consisting of safe Rust, dramatically decreasing the surface area where memory errors could originate.
